OptiFit: a fast method for fitting amplicon sequences to existing OTUs

### 2021-01-29 paper/paper.Rmd
More details

Closed-reference methods assign sequences to a set of pre-made OTUs generated
from reference sequences. If a query sequence is not within the distance
threshold to any of the reference sequences, it is discarded. _De novo_ methods
cluster sequences based on their distance to each other, without the use of an
external reference. Open-reference methods first perform closed-reference
clustering, then any sequences that cannot be assigned to reference OTUs are
clustered _de novo_ to create additional OTUs. While reference-based clustering
is generally fast, it is limited by the diversity represented in the reference
database. _De novo_ clustering overcomes the limitations of reference databases
by considering only sequences in the dataset, but is more computationally
intensive and generates different OTU assignments when new sequences are
introduced.

@@

To evaluate the OptiFit algorithm and compare to existing methods, we used four
published datasets isolated from soil [@johnston_metagenomics_2016], marine
[@henson_artificial_2016], mouse gut [@schloss_stabilization_2012], and human
gut [@baxter_microbiota-based_2016] samples. There are two strategies for
generating OTUs with OptiFit: 1) fit sequences to reference OTUs of an
independent database, or 2) split the dataset into a reference and query
fraction, then fit the query sequences to OTUs generated by clustering the
reference sequences _de novo_. For each dataset repeated with 100 random seeds,
we generated OTUs with OptiFit using both strategies, and also clustered _de
novo_ OTUs with OptiClust for comparison. All clustering was performed at a
sequence dissimilarity threshold of 0.03 and OTU quality was evaluated using the
Matthews Correlation Coefficient (MCC) as described previously
[@schloss_application_2016; @westcott_novo_2015]. We calculated the fraction of
query sequences that were fit to existing OTUs in closed-reference mode as an
additional measure of quality for this mode.

@@

To evaluate reference-based clustering with independent databases, we fit each
dataset to reference OTUs generated by _de novo_ clustering the Greengenes
database (v13_8_99), Silva non-redundant database (v132), and the Ribosomal
Database Project (RDP; v16). In open-reference mode, OTU quality was similar
between fitting the datasets to reference OTUs with OptiFit and clustering the
datasets _de novo_ with OptiClust. This held true for all datasets and reference
databases. However, in closed-reference mode, OTU quality was slightly worse
when fitting to greengenes and silva, and much worse when fitting to RDP as
compared to OptiClust. No more than half of query sequences were fit to
reference OTUs in closed-reference mode across any dataset/database combination.
**TODO: specify exact numbers for fraction mapped.**
OptiFit was able to fit more query sequences to reference OTUs created with the
greengenes and silva databases than with RDP. In terms of runtime,
closed-reference OptiFit outperformed OptiClust, while OptiClust outperformed
open-reference OptiFit.

To compare to existing software, vsearch was used to cluster OTUs _de novo_ or
with reference-based clustering to the greengenes database. For all datasets and
clustering methods (_de novo_, open reference, and closed reference), mothur's
clustering algorithms produced higher quality OTUs than vsearch. When
closed-reference clustering against the greengenes database, vsearch was able to
map more query sequences to the reference than mothur's OptiFit algorithm. In
terms of runtime, OptiFit generally performed faster than vsearch when reference
clustering, while vsearch _de novo_ clustering outperformed OptiClust.

@@

OTU quality from the split dataset strategy with OptiFit was highly similar to
that from _de novo_ clustering the whole dataset with OptiClust regardless of
mode. OTU quality was remarkably stable across 100 different random seeds. In
terms of runtime, closed-reference OptiFit performed faster than OptiClust on
whole datasets. In open-reference mode, OptiClust performed faster than OptiFit
only when the OptiFit reference fraction was 30% or less. The split dataset
strategy performed just as well as the database strategy in open-reference mode
regardless of database used, and outperformed the database strategy in
closed-reference mode.

We also tested three methods for selecting the sequences to be used as the
reference; a simple random sample, weighting sequences by relative abundance,
and weighting by similarity to other sequences in the dataset. OTU quality was
similar with the simple and abundance-weighted sampling, but slightly worse with
similarity-weighted sampling. In closed-reference mode, The fraction of query
sequences that can be fit to the reference OTUs decreases as the reference
fraction increases.

@@

We developed a new algorithm for fitting sequences to existing OTUs and have
demonstrated its suitability for reference-based clustering. OptiFit makes the
iterative method employed by OptiClust available for tasks where
reference-based clustering is required. We have shown that OTU quality is
similar between OptiClust and OptiFit in open-reference mode, regardless of
strategy employed. Open-reference OptiFit does perform slower than OptiClust
due to the additional _de novo_ clustering step, so users may prefer OptiClust
for tasks that do not require reference OTUs.

When fitting to public databases, OTU quality dropped in closed-reference mode
to different degrees depending on the database and dataset source, and no more
than half of query sequences were able to be fit to OTUs across any
dataset/database combination. This may reflect limitations of reference
databases, which are unlikely to contain sequences from rare and novel microbes.
This drop in quality was most notable with RDP. We recommend users who require
an independent reference database opt for Greengenes or Silva instead. Since
OptiClust performs faster than open-reference OptiFit and creates higher quality
OTUs than closed-reference OptiFit with the database strategy, we recommend
using OptiClust rather than fitting to a database where possible.
(**Pat: if you don't have breadth, closed-ref will suck.**)

The mothur algorithms produced higher quality OTUs than vsearch in
open-reference, closed-reference, or _de novo_ modes. However, vsearch was able
to fit more sequence into OTUs than OptiFit in closed-reference mode. While both
mothur and vsearch use a dissimilarity threshold for determining how to assign
sequences into OTUs, vsearch is more permissive than mothur. Mothur requires
that all pairs of sequences in an OTU are within the dissimilarity threshold
without penalizing the MCC, while vsearch only requires sequences to be similar
to one other sequence in the OTU. In this way, vsearch sacrifices OTU quality in
order to allow more sequences to fit to OTUs. Users who require closed-reference
clustering may prefer to use vsearch if they wish to maximize the fraction of
sequences that can be fit at the cost of OTU quality. However, mothur's
OptiClust or OptiFit are preferred for _de novo_ or open-reference clustering.

>>>>

There are three main categories of OTU clustering algorithms: closed reference,
open reference, and _de novo_ clustering. Closed reference methods assign
sequences to a set of pre-made OTUs generated from reference sequences. If a
query sequence is not within the distance threshold to any of the reference
sequences, it is discarded. While reference-based clustering is generally fast,
it is limited by the diversity represented in the reference database. _De novo_
methods cluster sequences based on their distance to each other, without the use
of an external reference. open reference methods first perform closed reference
clustering, then any sequences that cannot be assigned to reference OTUs are
clustered _de novo_ to create additional OTUs.  _De novo_ clustering overcomes
the limitations of reference databases by considering only sequences in the
dataset, but is more computationally intensive and generates different OTU
assignments when new sequences are introduced. Previous studies found that _de
novo_ clustering created the highest quality OTU assignments based on the
Matthews correlation coefficient (MCC) [@schloss_application_2016;
@westcott_novo_2015].

@@

To evaluate the OptiFit algorithm and compare to existing methods, we used four
published datasets isolated from soil [@johnston_metagenomics_2016], marine
[@henson_artificial_2016], mouse gut [@schloss_stabilization_2012], and human
gut [@baxter_microbiota-based_2016] samples. There are two strategies for
generating OTUs with OptiFit: 1) fit sequences to reference OTUs of an
independent database, or 2) split the dataset into a reference and query
fraction, then fit the query sequences to OTUs generated by clustering the
reference sequences _de novo_. For each dataset repeated with 100 random seeds,
we generated OTUs with OptiFit using both strategies, and also clustered _de
novo_ OTUs with OptiClust for comparison. All clustering was performed at a
sequence dissimilarity threshold of 0.03 and OTU quality was evaluated using the
Matthews Correlation Coefficient (MCC) as described previously
[@schloss_application_2016; @westcott_novo_2015]. We calculated the fraction of
query sequences that were fit to existing OTUs in closed reference mode as an
additional measure of quality for this mode.

@@

To evaluate reference-based clustering with independent databases, we fit each
dataset to reference OTUs generated by _de novo_ clustering the Greengenes
database (v13_8_99), Silva non-redundant database (v132), and the Ribosomal
Database Project (RDP; v16). In open reference mode, OTU quality was similar
between fitting the datasets to reference OTUs with OptiFit and clustering the
datasets _de novo_ with OptiClust. This held true for all datasets and reference
databases. However, in closed reference mode, OTU quality was slightly worse
when fitting to greengenes and silva, and much worse when fitting to RDP as
compared to OptiClust. No more than half of query sequences were fit to
reference OTUs in closed reference mode across any dataset/database combination.
**TODO: specify exact numbers for fraction mapped.**
OptiFit was able to fit more query sequences to reference OTUs created with the
greengenes and silva databases than with RDP. In terms of runtime,
closed reference OptiFit outperformed OptiClust, while OptiClust outperformed
open reference OptiFit.

To compare to existing software, vsearch was used to cluster OTUs _de novo_ or
with reference-based clustering to the greengenes database. For all datasets and
clustering methods (_de novo_, open reference, and closed reference), mothur's
clustering algorithms produced higher quality OTUs than vsearch. When
closed reference clustering against the greengenes database, vsearch was able to
map more query sequences to the reference than mothur's OptiFit algorithm. In
terms of runtime, OptiFit generally performed faster than vsearch when reference
clustering, while vsearch _de novo_ clustering outperformed OptiClust.

@@

OTU quality from the split dataset strategy with OptiFit was highly similar to
that from _de novo_ clustering the whole dataset with OptiClust regardless of
mode. OTU quality was remarkably stable across 100 different random seeds. In
terms of runtime, closed reference OptiFit performed faster than OptiClust on
whole datasets. In open reference mode, OptiClust performed faster than OptiFit
only when the OptiFit reference fraction was 30% or less. The split dataset
strategy performed just as well as the database strategy in open reference mode
regardless of database used, and outperformed the database strategy in
closed reference mode.

We also tested three methods for selecting the sequences to be used as the
reference; a simple random sample, weighting sequences by relative abundance,
and weighting by similarity to other sequences in the dataset. OTU quality was
similar with the simple and abundance-weighted sampling, but slightly worse with
similarity-weighted sampling. In closed reference mode, The fraction of query
sequences that can be fit to the reference OTUs decreases as the reference
fraction increases.

@@

We developed a new algorithm for fitting sequences to existing OTUs and have
demonstrated its suitability for reference-based clustering. OptiFit makes the
iterative method employed by OptiClust available for tasks where
reference-based clustering is required. We have shown that OTU quality is
similar between OptiClust and OptiFit in open reference mode, regardless of
strategy employed. open reference OptiFit does perform slower than OptiClust
due to the additional _de novo_ clustering step, so users may prefer OptiClust
for tasks that do not require reference OTUs.

When fitting to public databases, OTU quality dropped in closed reference mode
to different degrees depending on the database and dataset source, and no more
than half of query sequences were able to be fit to OTUs across any
dataset/database combination. This may reflect limitations of reference
databases, which are unlikely to contain sequences from rare and novel microbes.
This drop in quality was most notable with RDP. We recommend users who require
an independent reference database opt for Greengenes or Silva instead. Since
OptiClust performs faster than open reference OptiFit and creates higher quality
OTUs than closed reference OptiFit with the database strategy, we recommend
using OptiClust rather than fitting to a database where possible.
(**TODO: "if you don't have breadth, closed ref will suck." - Pat**)

The mothur algorithms produced higher quality OTUs than vsearch in
open reference, closed reference, or _de novo_ modes. However, vsearch was able
to fit more sequence into OTUs than OptiFit in closed reference mode. While both
mothur and vsearch use a dissimilarity threshold for determining how to assign
sequences into OTUs, vsearch is more permissive than mothur. Mothur requires
that all pairs of sequences in an OTU are within the dissimilarity threshold
without penalizing the MCC, while vsearch only requires sequences to be similar
to one other sequence in the OTU. In this way, vsearch sacrifices OTU quality in
order to allow more sequences to fit to OTUs. Users who require closed reference
clustering may prefer to use vsearch if they wish to maximize the fraction of
sequences that can be fit at the cost of OTU quality. However, mothur's
OptiClust or OptiFit are preferred for _de novo_ or open reference clustering.
